A comparison of design-based and model-based approaches for finite population spatial data.

### 2021-12-04 inst/manuscript/manuscript.Rmd
background edits

The design-based and model-based approaches incorporate randomness in fundamentally different ways. In this section, we describe the role of randomness and its effects on subsequent inferences. We then discuss specific inference methods of the approaches for spatial data.

@@

The design-based approach assumes the population is fixed. Randomness is incorporated via the selection of units in a smapling frame according to a sampling design. A sampling frame is the set of all units available to be sampled. A sampling design assigns a positive probability of inclusion (inclusion probability) to each unit in the sampling frame. Some examples of commonly used sampling designs include simple random sampling, stratified random sampling, and cluster sampling. If a sampling design selects units from the sampling frame while ignoring their spatial locations, we call them "Independent Random Sampling" (IRS) designs. If a sampling design selects units from the sampling frame while incorporating their spatial locations, we call them spatially balanced designs. Spatially balanced designs can be obtained using the Generalized Random Tessellation Stratified (GRTS) algorithm [@stevens2004spatially], which we discuss in more detail in Section \ref{subsec:spb_design}. The design-based approach combines the randomness of the sampling design and the data collected via the sample to estimate fixed, unknown parameters (e.g., means and totals) of a population.

Treating the data as fixed and incorporating randomness through the sampling design yields estimators having very few other assumptions. Confidence intervals for these types of estimators are typically derived using limiting arguments that incorporate all possible randomizations of sampling units selected via the sampling design. Means and totals, for example, are asymptotically normally distributed (normal) by the Central Limit Theorem (under some assumptions). If we repeatedly sample the surface, then 95% of all 95% confidence intervals constructed from a procedure with appropriate coverage will contain the true, fixed mean. @sarndal2003model and @lohr2009sampling provide thorough reviews of the design-based approach.

The model-based approach assumes the data are a random realization of a data-generating stochastic process. Randomness is incorporated through distributional assumptions on this process. Strictly speaking, randomness need not be incorporated through random sampling, though @diggle2010geostatistical warn against preferential sampling. Preferential sampling occurs when the process generating the data locations and the process being modeled are not independent of one another. To guard against preferential sampling, model-based approaches often still implement random sampling.

Instead of estimating fixed but unknown parameters like a mean or total (as in the design-based approach), the goal of model-based inference in the spatial context is often to predict a realized variable, or value. For example, suppose the realized mean of all population units is the value of interest. Instead of \emph{estimating} a fixed, unknown mean, we are \emph{predicting} the value of the mean, a random variable. Prediction intervals are then derived using assumptions of the data generating process. If we repeatedly generate the response values from the same spatial process and sample, then 95% of all 95% prediction intervals constructed from a procedure with appropriate coverage will contain their respective realized means. @cressie1993statistics and @schabenberger2017statistical provide reviews of model-based approaches for spatial data. A visual comparison of the design-based and model-based assumptions is provided in Figure \ref{fig:fig1} (@verhoef2002sampling and @brus2021statistical provide similar figures).

```{r fig1, out.width = "100%", fig.cap = "A comparison of sampling under the design-based and model-based frameworks. Points circled are those that are sampled. In the top row, we have one fixed population, and three random samples of size four. The response values at each site are fixed, but we obtain different estimates for the mean response because the randomly sampled sites vary from sample to sample. In the bottom row, we have three realizations of the same spatial process sampled at the same locations. The spatial process generating the response values has a single mean, but the realized mean is different in each of the three panels."}
set.seed(10102021)

@@

The design-based approach can be used to select samples that are "well-spread" in space, or spatially balanced. Spatially balanced samples are useful because parameter estimates from these samples tend to vary less than parameter estimates from samples that are not spatially balanced [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. The first spatially balanced sampling algorithm that saw widespread use was the Generalized Random Tessellation Stratified (GRTS) algorithm [@stevens2004spatially]. To quantify the spatial balance of a sample, @stevens2004spatially proposed loss metrics based on Voroni polygons. After the GRTS algorithm was developed, several other spatially balanced sampling algorithms have emerged, including the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance Sampling [@benedetti2017spatially], and Halton Iterative Partitioning Sampling [@robertson2018halton]. In this manuscript, we use the Generalized Random Tessellation Stratified (GRTS) algorithm to select spatially balanced samples sampling because the algorithm has several attractive properties. It accommodates finite and infinite sampling frames, equal, unequal, and proportional (to size) inclusion probabilities, legacy (historical) sampling [@foster2017spatially], a minimum distance between units in a sample, and replacement units in a sample (which are units that can be sampled in place of an original unit that can no longer be sampled).
The GRTS algorithm samples from finite and infinite populations by utilizing a mapping between two-dimensional and one-dimensional space. The units in the two-dimensional sampling frame are divided into cells using a hierarchical address. This hierarchical address is then used to map the units from two-dimensional space to a one-dimensional line where each unit's line length equals its inclusion probability. A systematic sample is conducted on the line and linked back to a unit in two-dimensional space, which results in the desired sample. @stevens2004spatially provides further details.

@@

After selecting a spatially balanced sample using the GRTS algorithm (i.e., a GRTS sample), data are collected and used to estimate population parameters. To unbiasedly estimate population means and totals from sample data, one can use the Horvitz-Thompson estimator [@horvitz1952generalization]. If $\tau$ is a population total, the Horvitz-Thompson estimate of $\tau$, denoted by $\hat{\tau}_{ht}$, is is given by
\begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align}
where $Z_i$ is the value of the $i$th unit in the sample and $\pi_i$ is the inclusion probability of the $i$th unit in the sample. An estimate of the population mean can be obtained by dividing $\hat{\tau}_{ht}$ by the number of population units, $N$.

While the Horvitz-Thompson estimator is unbiased for population means and totals, it is also important to quantify the uncertainty in these estimates. @horvitz1952generalization and @sen1953estimate provide variance estimators for $\hat{\tau}_{ht}$, but these estimators have two drawbacks. First, they rely on calculating $\pi_{ij}$, the probability that unit $i$ and unit $j$ are both in the sample -- this quantity can be challenging if not impossible to calculate analytically. Second, these estimators ignore the spatial locations of the units in the sampling frame. To address these two drawbacks simultaneously, @stevens2003variance proposed the local neighborhood variance estimator. The local neighborhood variance estimator does not rely on $\pi_{ij}$ and incorporates spatial locations -- for technical details see @stevens2003variance. @stevens2003variance show the local neighborhood variance estimator tends to reduce the estimated variance of $\hat{\tau}$ compared to variance estimators ignoring spatial locations, yielding narrower confidence intervals for $\tau$.

@@

Finite Population Block Kriging (FPBK) is a model-based approach that expands the geostatistical Kriging framework to the finite population setting [@verhoef2008spatial]. Instead of developing inference based on a specific sampling design, we assume the data are generated by a spatial process. @verhoef2008spatial gives details on the theory of FPBK, but some of the basic principles are summarized below. Let ${\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}$ be an $N \times 1$ response vector at locations $s_1$, $s_2$, . . . , $s_N$ that can be measured at the $N$ population units. Suppose we want to predict some linear function of the response variable, $f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}$, where $\mathbf{b}^\prime$ is a $1 \times N$ vector of weights. For example, if we want to predict the population total across all population units, then we would use a vector of 1's for the weights.

We often only have a sample of the $N$ population units. Denoting quantities that are part of the sampled population units with a subscript \emph{s} and quantities that are part of the unsampled population units with subscript \emph{u}, let

\begin{equation}
\begin{pmatrix} \label{equation:Zmarginal}
    \mathbf{z}_s      \\
    \mathbf{z}_u
\end{pmatrix}
=
\begin{pmatrix}
  \mathbf{X}_s    \\
  \mathbf{X}_u
\end{pmatrix}
\bm{\beta} +
\begin{pmatrix}
\bm{\delta}_s    \\
\bm{\delta}_u
\end{pmatrix},
\end{equation}
where $\mathbf{X}_s$ and $\mathbf{X}_u$ are the design matrices for the sampled and unsampled population units, respectively, and $\bm{\beta}$ is the parameter vector of fixed effects.

Let $\bm{\delta} \equiv [\bm{\delta}_s \,\, \bm{\delta}_u]'$, where $\bm{\delta}_s$ and $\bm{\delta}_u$ are random errors for the sampled and unsampled population units, respectively. We assume $E(\bm{\delta}) = \mathbf{0}$ and that there is spatial correlation in $\bm{\delta}$ that can be modeled using a covariance function. It is common to assume the covariance function is second-order stationary and isotropic [@cressie1993statistics], and that the spatial covariance decreases as the separation between population units increases. Many spatial covariance functions exist, but the primary function we use throughout the simulations and applications in this manuscript is the exponential covariance function: the $i,j$th elment of the matrix $\cov(\bm{\delta})$ is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{1}$ is dependent random error variance measuring coarse-scale (correlated) variability, $\sigma^2_{2}$ is the independent random error variance measuring fine-scale (independent) variability, $\phi$ is the range parameter measuring the distance-decay rate of the correlation, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. Often $\sigma^2_{1}$ and $\sigma^2_{2}$ are called the partial sill and nugget, respectively. Any spatial covariance function could be used in the place of the exponential, including functions that allow for non-stationarity or anisotropy [@chiles1999geostatistics, pp. 80-93].

@@

We note that we only use FPBK in this paper in order to focus more on comparing the design-based and model-based approaches. Other methods, such as k-nearest-neighbors [@fix1989discriminatory; @ver2013comparison], random forest [@breiman2001random], Bayesian models [@chan2020bayesian], among others, could also be used to obtain predictions for a mean or total from spatially correlated responses of a finite population. We choose to use FPBK because it is faster than a Bayesian approach and it was developed with theoretically-based variance estimators of means and totals for spatial data, whereas random forests and k-nearest-neighbors use ad-hoc variance estimators in most cases [@ver2013comparison]; additionally, FBPK outperformed the other methods in most scenarios.

>>>>

The design-based and model-based approaches incorporate randomness in fundamentally different ways. In this section, we describe the role of randomness for each approach and the subsequent effects on statistical inferences for spatial data.

@@

The design-based approach assumes the population is fixed. Randomness is incorporated via the selection of units in a sampling frame. A sampling frame is the set of all units available to be sampled. Units from the sampling frame are selected as part of the sample according to a sampling design, which assigns a positive probability of inclusion (inclusion probability) to each unit from the sampling frame. Some examples of commonly used sampling designs include simple random sampling, stratified random sampling, and cluster sampling. When sampling designs incorporate spatial locations into sampling, we call the resulting samples "spatially balanced." One approach to selecting spatially balanced samples is the Generalized Random Tessellation Stratified (GRTS) algorithm [@stevens2004spatially], which we discuss in more detail in Section \ref{subsec:spb_design}. When sampling designs do not incorporate spatial locations into sampling, we call the resulting samples "non-spatially balanced."

Fundamentally, the design-based approach combines the randomness of the sampling design with the data collected via the sample to justify the estimation and uncertainty quantification of fixed, unknown parameters of a population (e.g., a population mean). Treating the data as fixed and incorporating randomness through the sampling design yields estimators having very few other assumptions. Confidence intervals for these types of estimators are typically derived using limiting arguments that incorporate all possible samples. Sample means, for example, are asymptotically normal (Gaussian) by the Central Limit Theorem (under some assumptions). If we repeatedly select samples from the population, then 95% of all 95% confidence intervals constructed from a procedure with appropriate coverage will contain the true, fixed mean. @sarndal2003model and @lohr2009sampling provide thorough reviews of the design-based approach.

The model-based approach assumes the data are a random realization of a data-generating stochastic process. Randomness is incorporated through distributional assumptions on this process. Strictly speaking, randomness need not be incorporated through random sampling, though @diggle2010geostatistical warn against preferential sampling. Preferential sampling occurs when the process generating the data locations and the process being modeled are not independent of one another. To guard against preferential sampling, model-based approaches often still implement some form of random sampling.

Instead of estimating fixed, unknown population parameters, as in the design-based approach, often the goal of model-based inference is to predict a realized variable, or value. For example, suppose the realized mean of all population units is the value of interest. Instead of \emph{estimating} a fixed, unknown mean, we are \emph{predicting} the value of the mean, a random variable. Prediction intervals are then derived using assumptions of the data-generating stochastic process. If we repeatedly generate response values from the same data-generating stochastic process and select samples, then 95% of all 95% prediction intervals constructed from a procedure with appropriate coverage will contain their respective realized means. @cressie1993statistics and @schabenberger2017statistical provide thorough reviews of model-based approaches for spatial data. In Figure \ref{fig:fig1}, we provide a visual comparison of the design-based and model-based approaches (@verhoef2002sampling and @brus2021statistical provide similar figures).

```{r fig1, out.width = "100%", fig.cap = "A visual comparison of the design-based and model-based approaches. In the top row, there is one fixed population with nine population units and three random samples of size four (points circled are those sampled). The response values at each site are fixed, but we obtain different estimates for the mean response in each random sample. In the bottom row, there are three realizations of the same data-generating stochastic process that are all sampled at the same four locations. The ata-generating stochastic process has a single mean, but the mean of the nine population units is different in each of the three realizations."}
set.seed(10102021)

@@

We previously mentioned that the design-based approach can be used to select spatially balanced samples (samples that incorporate spatial locations of the population units and are "well-spread" is space). Spatially balanced samples are useful because parameter estimates from these samples tend to vary less than parameter estimates from samples that are not spatially balanced [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. The first spatially balanced sampling algorithm seeing widespread use is the Generalized Random Tessellation Stratified (GRTS) algorithm [@stevens2004spatially]. To quantify the spatial balance of a sample, @stevens2004spatially proposed loss metrics based on Voronoi polygons (Direchlet Tessellations). After the GRTS algorithm was developed, several other spatially balanced sampling algorithms emerged, such as the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance Sampling [@benedetti2017spatially], and Halton Iterative Partitioning Sampling [@robertson2018halton]. In this manuscript, we select spatially balanced samples using the Generalized Random Tessellation Stratified (GRTS) algorithm because it has several attractive properties. More specifically, the GRTS algorithm accommodates finite and infinite sampling frames, equal, unequal, and proportional (to size) inclusion probabilities, legacy (historical) sampling [@foster2017spatially], a minimum distance between units in a sample, and replacement units (replacement units are population units that can be sampled when a population unit originally selected can no longer be sampled). The GRTS algorithm selects samples by utilizing a particular mapping between two-dimensional and one-dimensional space that preserves proximity relationships. Via this mapping, units in two-dimensional space are partitioned using a hierarchical address. This hierarchical address is used to map population units to a one-dimensional line. On the one dimensional line, each population unit's line length equals its inclusion probability. Then, a systematic sample of population units is selected on the line, yielding desired sample. @stevens2004spatially provides more technical details.

@@

After selecting a sample and collecting data, unbiased estimates of population means and totals can be obtained using the Horvitz-Thompson estimator [@horvitz1952generalization]. If $\tau$ is a population total, the Horvitz-Thompson estimate of $\tau$, denoted by $\hat{\tau}_{ht}$, is is given by
\begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align}
where $Z_i$ is the value of the $i$th population unit in the sample and $\pi_i$ is the inclusion probability of the $i$th population unit in the sample. An estimate of the population mean is obtained by dividing $\hat{\tau}_{ht}$ by $N$, the number of population units.

It is also important to quantify uncertainty $\hat{\tau}_{ht}$.  @horvitz1952generalization and @sen1953estimate provide variance estimators for $\hat{\tau}_{ht}$, but these estimators have two drawbacks. First, they rely on calculating $\pi_{ij}$, the probability that population unit $i$ and populatoin unit $j$ are both in the sample -- this quantity can be challenging if not impossible to calculate analytically. Second, these estimators ignore the spatial locations of the population units. To address these two drawbacks simultaneously, @stevens2003variance proposed the local neighborhood variance estimator. The local neighborhood variance estimator does not rely on $\pi_{ij}$ and incorporates spatial locations -- for technical details see @stevens2003variance. @stevens2003variance show the local neighborhood variance estimator tends to reduce the estimated variance of $\hat{\tau}$ and yield narrower confidence intervals compared to variance estimators that ignore spatial locations.

@@

Finite Population Block Kriging (FPBK) is a model-based approach that expands the geostatistical Kriging framework to the finite population setting [@verhoef2008spatial]. Instead of developing inference based on a specific sampling design, we assume the data are generated by a spatial stochastic process. We summarize some of the basic principles of FBPK next (for more technical details, see @verhoef2008spatial) Let ${\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}$ be an $N \times 1$ response vector at locations $s_1$, $s_2$, . . . , $s_N$ that can be measured at the $N$ population units. Suppose we want to use a sample to predict some linear function of the response variable, $f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}$, where $\mathbf{b}^\prime$ is a $1 \times N$ vector of weights (e.g, the population mean is represented by a weights vector whose elements all equal one). Denoting quantities that are part of the sampled population units with a subscript \emph{s} and quantities that are part of the unsampled population units with subscript \emph{u}, let

\begin{equation}
\begin{pmatrix} \label{equation:Zmarginal}
    \mathbf{z}_s      \\
    \mathbf{z}_u
\end{pmatrix}
=
\begin{pmatrix}
  \mathbf{X}_s    \\
  \mathbf{X}_u
\end{pmatrix}
\bm{\beta} +
\begin{pmatrix}
\bm{\delta}_s    \\
\bm{\delta}_u
\end{pmatrix},
\end{equation}
where $\mathbf{X}_s$ and $\mathbf{X}_u$ are the design matrices for the sampled and unsampled population units, respectively, $\bm{\beta}$ is the parameter vector of fixed effects, and $\bm{\delta} \equiv [\bm{\delta}_s \,\, \bm{\delta}_u]'$, where $\bm{\delta}_s$ and $\bm{\delta}_u$ are random errors for the sampled and unsampled population units, respectively.

FBPK assumes $\bm{\delta}$ in Equation$~$\ref{equation:Zmarginal} has mean zero and a spatial correlation structure that can be modeled using a covariance function. This covariance function is commonly assumed to be non-negative (between zero and one), second-order stationary (depending only on the distance between population units), isotropic (independent of direction), and decay with distance between population units [@cressie1993statistics]. Henceforth, it is implied that we have made these same assumptions regarding $\bm{\delta}$, though @chiles1999geostatistics, pp. 80-93 discuss covariance functions that are not second-order stationary, not isotropic, or both. A variety of flexible covariance functions can be used to model $\bm{\delta}$ [@cressie1993statistics]; one example is the exponential covariance function (for a thorough list of spatial covariance functions, see @cressie1993statistics. The $i,j$th element of the exponential covariance matrix, $\cov(\bm{\delta})$, is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{1}$ is the variance parameter quantifying the variability that is dependent (coarse-scale), $\sigma^2_{2}$ is the variance parameter quantifying the variability that is independent (fine-scale), $\phi$ is the range parameter measuring the distance-decay rate of the covariance, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. The proportion of variability attributable to dependent random error is $\sigma^2_{1} / (\sigma^2_{1} + \sigma^2_{2})$. Similarly, the proportion of variability attributable to independent random error is $\sigma^2_{2} / (\sigma^2_{1} + \sigma^2_{2})$. Finally we note that $\sigma^2_{1}$ and $\sigma^2_{2}$ are often called the partial sill and nugget, respectively.

@@

Other approaches, such as k-nearest-neighbors [@fix1989discriminatory; @ver2013comparison], random forests [@breiman2001random], Bayesian models [@chan2020bayesian], among others, could also be used to obtain predictions for a mean or total from spatially correlated responses of a finite population. Compared to the k-nearest-neighbors and random forest approach, we prefer FBPK because it is model-based and relies on theoretically-based variance estimators leveraging the model's spatial covariance structure, whereas k-nearest-neighbors and random forests use ad-hoc variance estimators [@ver2013comparison]. Additionally, @ver2013comparison studied compared FBPK, k-nearest-neighbors, and random forests in a variety of spatial data contexts, and FBPK tended to perform best. Compared to the Bayesian approach, we prefer FPBK mostly because it is much more computationally efficient.
